---
title: "Samples for protein assays"
output: html_document
date: "2024-09-30"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load libraries, echo=FALSE, message=FALSE, warning = FALSE}

library(ggplot2)
source("DYNAMHIX_Functions.R")

```


## Overview

This doc just describes choosing samples for batches for serum protein assays. Includes:

* Zonulin
* ALP

For the first batch (and this will initially be tried on zonulin kits) we will make sure that there are >2 aliquots of serum available for use and there are 5 timepoints available. 

## Checking the samples that we have

These are the samples that are either PSC-UC or UC.

```{r samples, echo=FALSE, message=FALSE}

dat <- read.csv("CMS031NIIL_DATA_LABELS_2024-09-30_0800.csv",
                header = TRUE,
                stringsAsFactors = FALSE,
                sep="\t")

# recode the variables
dat <- recode_dat(dat)

# all of the aliquot data that we have
aliquots <- get_aliquot_data(dat)

# get patient ids when UC or PSC-UC
diagnosis_to_keep <- dat$Participant.ID[dat$Diagnosis == "PSC-UC" | dat$Diagnosis == "UC"]

# aliquots to keep based on PSC-UC or UC diagnosis
aliquots <- aliquots[aliquots$Patient_ID %in% diagnosis_to_keep,]

# just get the serum samples
aliquots <- aliquots[grep("_SE_", aliquots$Sample_ID),]

# create kitID for aliquots
kitid <- strsplit(aliquots$Sample_ID, "_")
kitid <- lapply(kitid, function(x) paste0(x[1], "_", x[2]))
kitid <- unlist(kitid)

aliquots$Kit_ID <- kitid
  
# get table of total numbers
df_aliquots <- data.frame(Total_Patients = length(unique(aliquots$Patient_ID)),
                          Total_Aliquots = nrow(aliquots))


```

## Number of aliquots per sample

Below is a table that shows the number of aliquots per sample (i.e. kit id).

```{r aliquots per sample, echo=FALSE, message=FALSE}

sample_aliquots <- as.data.frame(table(c(aliquots$Kit_ID)))
colnames(sample_aliquots) <- c("Sample", "No. Aliquots")

knitr::kable(sample_aliquots)

```

## Get timepoint information

Here I just look at the number of timepoints per patient.

```{r timepoints, echo=FALSE, message=FALSE}

timepoints <- get_variable_data(dat, variable = "Event.Name")
timepoints <- timepoints[timepoints$PID %in% diagnosis_to_keep,]

df_timepoints <- as.data.frame(table(timepoints$PID))
colnames(df_timepoints) <- c("Patient_ID", "No. Timepoints")

knitr::kable(df_timepoints)
```


## Get batches

The first batch is going to be the ones that have enough aliquots and 5 timepoints - this is because we are trying out the kits (ZO1).


```{r batches, echo=FALSE, message=FALSE}

# get patients with 5 timepoints
fivers <- df_timepoints[df_timepoints$`No. Timepoints` == 5,]

# get aliquots with 5 timepoints and max serum samples
samples <- sample_aliquots[sample_aliquots$`No. Aliquots` >= 3,]

# both criteria
aliquots_to_sample <- aliquots[aliquots$Patient_ID %in%
                                 fivers$Patient_ID &
                                 aliquots$Kit_ID %in%
                                 samples$Sample,]

# re-check the number of patient samples - should be five but not
# all neccesarily have a full set of aliquots
#patients_fit <- as.data.frame(table(aliquots_to_sample$Patient_ID))

aliquots_to_sample <- aliquots_to_sample[order(aliquots_to_sample$Patient_ID),]

# just the SE_A ones
aliquots_to_sample <- aliquots_to_sample[grep("SE_A", aliquots_to_sample$Sample_ID),]

# readjust based on patients that still have all 5
x <- as.data.frame(table(aliquots_to_sample$Patient_ID))
x <- x[order(x$Freq, decreasing=TRUE),]

# need 42 samples to start with - take the first 8 patients
patients_to_take <- as.character(x[1:8, 1])

# Also add two extras - going to add GI9105 first two timepoints

aliquots_to_take <- aliquots_to_sample[aliquots_to_sample$Patient_ID %in% patients_to_take |
                                         aliquots_to_sample$Sample_ID == "CMS031_030_SE_A" |
                                         aliquots_to_sample$Sample_ID == "CMS031_057_SE_A",]



# add the aliquot id as the rownames
rownames(aliquots_to_take) <- aliquots_to_take$Sample_ID

# do the same with the original aliquots file
rownames(aliquots) <- aliquots$Sample_ID

# add batch name to the aliquots table
aliquots$batch <- NA

aliquots$batch <- ifelse(aliquots$Sample_ID %in% aliquots_to_take$Sample_ID, 1, aliquots$batch)

## Final table for Kate to use
aliquots <- aliquots[order(aliquots$batch, aliquots$Patient_ID),]

final_aliquot_table <- data.frame(aliquot_id = aliquots$Sample_ID,
                                  kit_id = aliquots$Kit_ID,
                                  patient_id = aliquots$Patient_ID,
                                  box = aliquots$Box,
                                  box_position = aliquots$Position,
                                  batch = aliquots$batch)

write.table(final_aliquot_table,
            file = "DYNAMHIX_Aliquot_Batches.tsv",
            quote=FALSE,
            row.names=FALSE,
            sep="\t")

```









